Foundations of ecological and evolutionary change

### 2025-06-01 foundations_of_ecoevo.tex
Rewrite the introduction along reviewer recommendations; cut some things to the first section, and prepare for a better verbal explanation of eqn 1

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\citeproc{ref-Kitcher1993}{Kitcher 1993}),
and its value is perhaps most evident in reconciling unconnected models
and revealing new and unexpected empirical predictions. In theoretical
physics, Newton's unification of celestial and terrestrial motion, and
Maxwell's unification of electricity and magnetism, are among the most
important advances in scientific history. Likewise, in the life
sciences, the modern synthesis of natural selection and Mendelian
inheritance revolutionised how mechanisms underlying evolution and
genetics are understood and applied (\citeproc{ref-Gayon2019}{Gayon and
Huneman 2019}). The analogy to physics is imperfect because evolution
encompasses non-biological systems with non-genetic inheritance
(\citeproc{ref-Price1995}{Price 1995};
\citeproc{ref-Shennan2011}{Shennan 2011};
\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
\citeproc{ref-Lewens2015}{Lewens 2015}), but to illustrate our point, we
restrict our focus to biological evolution. Following the modern
synthesis, patterns in evolutionary biology and genetics could be
interpreted through a common framework.

Selection, mutation, recombination, migration, and drift all have
predictable consequences for both phenotypes and genotypes in a
population. These mechanisms are therefore foundational to understanding
evolution and genetics, and their roles in explaining evolutionary and
genetic phenomena are no more conceptually pluralistic than that of
gravity in explaining celestial and terrestrial motion. Instead,
biologists understand that each of these mechanisms will jointly affect
genetic composition and evolutionary change in a population. For
example, all else being equal, drift will lead to a loss of both
heterozygosity and adaptive phenotypes. This constrains the possible
relations that can exist between genetic composition and evolutionary
change and therefore clarifies the scope of hypotheses that are
biologically coherent (\citeproc{ref-Caswell1988}{Caswell 1988}). It
would obviously be mistaken to model a population in which drift had
inconsistent effects on evolution and genetics, but such mistakes would
surely be common without the foundational theory required to unify the
two subjects. The absence of this theory is almost unthinkable in modern
biology. Evolution and genetics are arguably as intertwined in biology
as electricity and magnetism are in physics.

In contrast, theory underlying ecology and evolution has not been so
tightly intertwined. Until recently, population ecology and evolution in
particular have taken almost parallel paths for developing theory.
Although many revelatory models exist to connect ecological and
evolutionary phenomena (\citeproc{ref-Lion2018}{Lion 2018};
\citeproc{ref-Patel2018}{Patel et al. 2018};
\citeproc{ref-Govaert2019}{Govaert et al. 2019}), the foundational
assumptions and equations of ecology and evolution in these models
remain mostly separated conceptually. But like the mechanisms driving
evolutionary and genetic change, those that drive ecological and
evolutionary change are well-established with predictable joint
consequences. In evolution, natural selection, drift, and gene flow are
all instantiated through a combination of different ecological factors
(\citeproc{ref-Connor2004}{Connor and Hartl 2004}). Most notably, change
in population density and phenotype are both caused by individual birth
and death (\citeproc{ref-Turchin2001}{Turchin 2001}). This should allow
biologists to place ecological and evolutionary change on a shared
theoretical foundation.

We believe that such a foundation exists for population ecology and
evolutionary biology, and that it can bridge the fundamental equations
of both subjects and serve as a starting point for understanding change
in any population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues for
a simple birth and death model (Box 1) as a fundamental equation of
population ecology, and for exponential growth to consequently be as
fundamental to ecology as Newton's first law of motion is to physics.
Turchin (\citeproc{ref-Turchin2001}{2001}) argues that general
principles are needed to establish a logical foundation for population
ecology. In evolutionary biology, the Price equation (Box 2) is almost
universally regarded as fundamental (\citeproc{ref-Price1970}{Price
1970}; \citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Gardner2008}{Gardner 2008};
\citeproc{ref-Frank2015a}{Frank 2017}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Luque2021}{Luque and Baravalle 2021}). It
exhaustively and exactly describes evolutionary change for any closed
evolving population (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Luque2016}{Luque 2017};
\citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}), and it serves as the
starting point for all other fundamental equations in evolution
(\citeproc{ref-Queller2017}{Queller 2017}). We believe that this logical
foundation can be made sufficiently broad to encompass both population
and evolutionary change.

We propose an equation from which the fundamental ecological and
evolutionary equations can be derived. Derivation follows by adding
assumptions that are specific to population ecology or evolution in the
same way that key equations of population genetics (e.g., average
excess) or quantitative genetics (e.g., Breeder's equation) can be
derived from the Price equation (\citeproc{ref-Queller2017}{Queller
2017}). We hope that our equation can provide a starting point for
building a shared logical framework between ecology and evolution.

@@

\begin{quote}
\textbf{Box 1:} The number of individuals in any closed population
(\(N\)) at any given time (\(t + 1\)) is determined by the existing
number of individuals (\(N_{t}\)), plus the number of births
(\(Births\)) minus the number of deaths (\(Deaths\)),
\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
for any closed population. Despite its simplicity, it is a general
equation for defining population change and a starting point for
understanding population ecology. Turchin
(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
fundamental equation is the tendency for populations to grow
exponentially (technically geometrically in the above case where time is
discrete). This inherent underlying tendency towards exponential growth
persists even as the complexities of real populations, such as
structure, stochasticity, or density-dependent effects are added to
population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
assumption that all individuals in the population are identical, a per
capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
size of any closed population equals its existing size times its finite
rate of increase.
\end{quote}

@@

\begin{quote}
\textbf{Box 2:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\citeproc{ref-Price1970}{Price 1970},
\citeproc{ref-Price1972}{1972}), it postulates some basic properties
that all evolutionary systems must satisfy: change over time, ancestor
and descendant relations, and a character or phenotype
(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
the Price equation represents evolutionary change with the predominant
notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
2012}). A more specific version of the covariance term was already known
within the quantitative and population genetics tradition
(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
action of natural selection. The Price equation adds an expectation term
and abstracts away from any specific mechanisms of replication or
reproduction, or mechanisms of inheritance. Its definitional nature and
lack of substantive biological assumptions has been portrayed both as a
strength (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
greatest weakness. The abstract nature of the Price equation places it
at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fisher's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
2017}). This abstractness is also key to developing a more general view
of evolution (\citeproc{ref-Rice2020}{Rice 2020};
\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
\citeproc{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

@@

To fully unify biological evolution and population ecology, we must
reconcile the general equation for population change (Box 1) with the
Price equation (Box 2). Previous theory has produced mathematical
synthesis between ecological and evolutionary equations
(\citeproc{ref-Page2002}{Page and Nowak 2002};
\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but not
established a unified conceptual framework for joint ecological and
evolutionary change. For example, Page and Nowak
(\citeproc{ref-Page2002}{2002}) demonstrate a mathematical equivalence
between the Price equation and a general equation for community
dynamics. In both equations, the relative frequencies of different
ecological (species) or evolutionary (phenotype) variants map to their
respective ecological (population growth) or evolutionary (fitness)
change in a way that is mathematically equivalent. Collins and Gardner
(\citeproc{ref-Collins2009}{2009}) use the Price equation to build a
general quantitative framework that partitions phenotypic change into
ecological, evolutionary, and physiological components. The average
change in a community-wide trait (e.g., carbon uptake) is exactly and
generally partitioned into change attributable to changes in species
composition, expected evolution of lineages within species, and expected
change in lineage physiology across all species. Both Page and Nowak
(\citeproc{ref-Page2002}{2002}) and Collins and Gardner
(\citeproc{ref-Collins2009}{2009}) use relative species abundances,
meaning that species frequencies must sum to one. This is useful for
demonstrating mathematical equivalence (\citeproc{ref-Page2002}{Page and
Nowak 2002}) or expected change in a community-wide trait
(\citeproc{ref-Collins2009}{Collins and Gardner 2009}), but a different
approach is needed to recover absolute change in population size and
thereby recover a joint foundation for ecology and evolution.

The Price equation is critical for partitioning different components of
biological change (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Gardner2008}{Gardner
2008}; \citeproc{ref-Luque2016}{Luque 2017};
\citeproc{ref-Queller2017}{Queller 2017};
\citeproc{ref-Lehtonen2018}{Lehtonen 2018}). It has also been highly
useful for integrating evolutionary theory across disciplines
(\citeproc{ref-Fox2006}{Fox 2006};
\citeproc{ref-Brantingham2007}{Brantingham 2007};
\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
\citeproc{ref-Frank2015}{Frank 2015}; \citeproc{ref-Godsoe2021}{Godsoe
et al. 2021}; \citeproc{ref-Ulrich2024}{Ulrich et al. 2024}). These
properties would seem to make it an intuitive starting point for a
logical foundation of ecology and evolution, perhaps through some kind
of mathematical equivalence (\citeproc{ref-Page2002}{Page and Nowak
2002}) or addition of terms (\citeproc{ref-Collins2009}{Collins and
Gardner 2009}), or through the use of its recursive structure
(\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\citeproc{ref-Frank2012}{Frank 2012}). But despite its flexibility, the
Price equation still relies on relative frequencies, which must by
definition sum to one (\citeproc{ref-Frank2015}{Frank 2015}). This is
because the Price equation describes the average change in a population;
the frequency of entities is scaled thereby conserving total probability
(\citeproc{ref-Frank2015}{Frank 2015}, \citeproc{ref-Frank2016}{2016}).
But to recover the fundamental principle of exponential population
growth (\citeproc{ref-Turchin2001}{Turchin 2001}), this scaling must be
avoided in a fundamental equation of ecology and evolution.

@@

From eqn 1, we can derive the most fundamental equations of population
ecology (Box 1) and evolutionary biology (Box 2) through an appropriate
interpretation of \(z_{i}\).

@@

To recover the general equation for population ecology (Box 1), we
define \(z_{i}\) as the identity of \(i\) belonging to the population.
In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
member of the population. Further, we assume that individuals do not
change species by setting \(\Delta z_{i} = 0\). In this case,

@@

If we define \(\lambda = 1 + \beta - \delta\) (Box 1), then we can
rewrite,

@@

From eqn 1, which describes fundamental birth and death processes in a
population, we can derive both the most fundamental model of population
ecology (eqn 2; Box 1) and the fundamental equation of evolution (eqn
10; Box 2). We might also consider \(\Omega\) to be a quantity of
ecosystem function, in that it is the summed effect of the traits
(\(z_{i}\)) of individual entities (\(i\)) in the population. For
example, \(z\) might be interpreted as biomass or photosynthetic rate.

@@

Rather than making simplifying assumptions, as is the approach for most
ecological and evolutionary models, we focus on fundamental axioms that
are universal to closed biological systems: discrete individuals, birth,
death, and change over time. We define an abstract sum (\(\Omega\)), to
which all individuals within the population contribute. From the simple
assumptions of population identity (\(z_{i} = 1\)) and invariability
(\(\Delta z_{i} = 0\)), we recover the most general equation of
population ecology (Box 1) and principle of exponential growth
(\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
and applying the total conservation of probability to individual
frequencies (\citeproc{ref-Frank2015}{Frank 2015},
\citeproc{ref-Frank2016}{2016}), we recover the most fundamental
equation of evolution (Box 2). We therefore deliver a general framework
to unify fundamental equations of ecological and evolutionary change.

@@

Our unification demonstrates the equivalence between the finite rate of
increase \(\lambda\) (Box 1) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
and population mean evolutionary fitness \(\bar{w}\) (Box 2). The
population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
rewritten as \(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first
to notice this equivalence between population growth and mean population
fitness. Indeed, the relationship between population growth rate and
evolutionary fitness has been proposed and applied many times before
(\citeproc{ref-Fisher1930}{Fisher 1930};
\citeproc{ref-Charlesworth1980}{Charlesworth 1980};
\citeproc{ref-Lande1982}{Lande 1982}; \citeproc{ref-Roff2008}{Roff
2008}; \citeproc{ref-Lion2018}{Lion 2018}). For example, Lande
(\citeproc{ref-Lande1982}{1982}) explicitly concludes mean absolute
fitness per unit time is \(\bar{w} = e^{r}\), where \(r = ln(\lambda)\).
We show this from first principles and clarify the relationship between
fitness and population growth. Over an arbitrary length of time, fitness
is properly defined as \(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an
individual's lifetime (which, by definition, includes death), fitness is
therefore \(\beta_{i}\). Interestingly, the rate of change in ecology
and evolution are reflected in the first and second statistical moments
of fitness, respectively. Population growth rate reflects mean fitness
\(\bar{w}\), while the rate of evolutionary change reflects the variance
in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem)
(\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
2004}; \citeproc{ref-Queller2017}{Queller 2017}).

@@

We have focused on the dynamics of a closed population, and in doing so
leave ecological and evolutionary change attributable to migration for
future work. In population ecology, immigration and emigration can be
incorporated by adding a term for each to the right-hand side of the
equation in Box 1 (\citeproc{ref-Gotelli2001}{Gotelli 2001}). In
evolution, because the Price equation relies on mapping
ancestor-descendant relationships, accounting for migration is more
challenging. Kerr and Godfrey-Smith (\citeproc{ref-Kerr2009}{2009})
demonstrate how the Price equation can be extended to allow for
arbitrary links between ancestors and descendants, thereby extending the
Price equation to allow for immigration and emigration. Frank
(\citeproc{ref-Frank2012}{2012}) presents a simplified version of Kerr
and Godfrey-Smith (\citeproc{ref-Kerr2009}{2009}) that allows some
fraction of descendants to be unconnected to ancestors. In both ecology
and evolution, accounting for migration is done through the use of
additional terms on the right-hand side of the equations.

>>>>

Theoretical unification of phenomena is a powerful tool for scientific
advancement. Such unification has been a major goal in scientific
research throughout history (\citeproc{ref-Kitcher1993}{Kitcher 1993}),
and its value is perhaps most evident in reconciling unconnected models
and revealing new and unexpected empirical predictions. In evolutionary
biology, the Price equation (Box 1) provides a unifying framework for
evolutionary theory by exhaustively and exactly describing evolutionary
change for any closed evolving population
(\citeproc{ref-Price1970}{Price 1970}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Lehtonen2020}{Lehtonen et al. 2020}). The Price
equation is therefore fundamental, in the sense that it binds together
all of evolutionary theory by formally defining what evolutionary change
is and is not (\citeproc{ref-Price1970}{Price 1970};
\citeproc{ref-Rice2004}{Rice 2004}; \citeproc{ref-Gardner2008}{Gardner
2008}; \citeproc{ref-Frank2015a}{Frank 2017};
\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Luque2021}{Luque and
Baravalle 2021}). Using this formal definition, the scope of, and
relationships among, sub-disciplines within evolutionary theory can be
clarified. For example, fundamental equations of both population and
quantitative genetics can be derived from the Price equation
(\citeproc{ref-Queller2017}{Queller 2017}). This provides conceptual
clarity by demonstrating the logical consistency of different
theoretical frameworks within evolutionary biology. Our aim here is to
propose an equation that extends this conceptual clarity to include
population change and thereby provide a formal and exact definition of
joint ecological and evolutionary change.

In biological populations, ecological change is caused by the same
processes of individual birth and death that cause changes in allele
frequencies and phenotypes (\citeproc{ref-Turchin2001}{Turchin 2001};
\citeproc{ref-Connor2004}{Connor and Hartl 2004};
\citeproc{ref-Barfield2011}{Barfield et al. 2011}). As with evolution, a
fundamental equation can exhaustively and exactly define population
change. Unlike the Price equation, this fundamental equation is perhaps
self-evident; population change is simply the addition of births and
subtraction of deaths from current population size (\(N_{t}\)) to
recover the new population size (\(N_{t+1}\); Box 2). By definition,
this relationship of \(N_{t+1} = N_{t} + Births - Deaths\) applies to
any closed population. Turchin (\citeproc{ref-Turchin2001}{2001}) argues
that general principles are needed to establish a logical foundation for
population ecology, and this simple birth and death model and the
consequences that logically follow from it (e.g., exponential population
growth) is fundamental to population ecology. Any unifying definition of
joint ecological and evolutionary change must be able to derive both the
Price equation and this birth and death model.

The interdependence of ecological and evolutionary processes has long
been recognised (e.g., \citeproc{ref-Darwin1859}{Darwin 1859};
\citeproc{ref-Fisher1958}{Fisher 1958};
\citeproc{ref-Pelletier2009}{Pelletier et al. 2009}), but the rise of
eco-evolutionary models, which incorporate both, is relatively recent
following a widespread recognition that ecology and evolution can happen
on similar timescales (\citeproc{ref-Govaert2019}{Govaert et al. 2019};
\citeproc{ref-Yamamichi2023}{Yamamichi et al. 2023}). Currently, a
universally recognised formal definition of eco-evolutionary change is
lacking, with some authors broadly interpreting ``eco-evolutionary
dynamics'' to allow for a separation of ecological and evolutionary
timescales (\citeproc{ref-Lion2023}{Lion et al. 2023}) and others
advocating for a more narrow interpretation in which no such separation
is permitted (\citeproc{ref-Bassar2021}{Bassar et al. 2021}). Bassar et
al. (\citeproc{ref-Bassar2021}{2021}) recognise two types of
eco-evolutionary models that follow from these interpretations. The
first type uses separate equations to model population change versus
evolutionary change, thereby allowing for any number of ecological,
evolutionary, or environmental feedbacks (e.g.,
\citeproc{ref-Lion2018}{Lion 2018}; \citeproc{ref-Patel2018}{Patel et
al. 2018}; \citeproc{ref-Lion2023}{Lion et al. 2023}). The second type
models population demographics as functions of quantitative traits, with
ecological and evolutionary change following from demographic processes
and trait distributions operating on the same timescale (e.g.,
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Simmonds2020}{Simmonds et al. 2020};
\citeproc{ref-Jaggi2024}{Jaggi et al. 2024}). Both model types can be
very general, but like all predictive models, they rely on simplifying
assumptions for tractability (\citeproc{ref-Levins1966}{Levins 1966};
\citeproc{ref-Luque2016}{Luque 2017}). These simplifying assumptions can
be grounded in the Price equation to demonstrate accuracy and logical
consistency when modelling evolutionary change (e.g.,
\citeproc{ref-Coulson2008}{Coulson and Tuljapurkar 2008};
\citeproc{ref-Barfield2011}{Barfield et al. 2011};
\citeproc{ref-Rees2016}{Rees and Ellner 2016};
\citeproc{ref-Lion2018}{Lion 2018}). For example, Barfield et al.
(\citeproc{ref-Barfield2011}{2011}) link their model back to Price
(\citeproc{ref-Price1970}{1970}), which they consider to be a
``universal law of evolution'', to place their conclusions concerning
stage structured evolution in the broader context of evolutionary
theory. The role of fundamental equations is therefore important for
unifying theory (\citeproc{ref-Luque2016}{Luque 2017}), and we believe
that a fundamental equation of eco-evolutionary change is needed.

We present an equation from which the fundamental equations of ecology
and evolutionary biology can be derived. Derivation follows by adding
assumptions that are specific to population ecology or evolution in the
same way that key equations of population genetics (e.g., average
excess) or quantitative genetics (e.g., Breeder's equation) can be
derived from the Price equation (\citeproc{ref-Queller2017}{Queller
2017}). We propose our equation as a formal definition of
eco-evolutionary change.

@@

\begin{quote}
\textbf{Box 1:} The Price equation is an abstract formula to represent
evolutionary change. Formulated originally in the early 1970s by George
Price (\citeproc{ref-Price1970}{Price 1970},
\citeproc{ref-Price1972}{1972}), it postulates some basic properties
that all evolutionary systems must satisfy: change over time, ancestor
and descendant relations, and a character or phenotype
(\citeproc{ref-Rice2004}{Rice 2004}). Using simple algebraic language,
the Price equation represents evolutionary change with the predominant
notation,
\[\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).\]
In the above equation, \(\Delta\bar{z}\) is the change in the average
character value \(z\) over a time step of arbitrary length, \(w\) is an
individual's absolute fitness, and \(\bar{w}\) average population
fitness. On the right-hand side of the equation, the first term is the
covariance between a character value \(z\) and fitness \(w\), which
reflects \(\bar{z}\) change attributable to differential survival and
reproduction. The second term is the expected value of \(\Delta z\),
which reflects the extent to which offspring deviate from parents in
\(z\) (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Okasha2006}{Okasha 2006}; \citeproc{ref-Frank2012}{Frank
2012}). A more specific version of the covariance term was already known
within the quantitative and population genetics tradition
(\citeproc{ref-Robertson1966}{Robertson 1966}), usually representing the
action of natural selection. The Price equation adds an expectation term
and abstracts away from any specific mechanisms of replication or
reproduction, or mechanisms of inheritance. Its definitional nature and
lack of substantive biological assumptions has been portrayed both as a
strength (\citeproc{ref-Rice2004}{Rice 2004};
\citeproc{ref-Frank2012}{Frank 2012}; \citeproc{ref-Luque2016}{Luque
2017}; \citeproc{ref-Baravalle2022}{Baravalle and Luque 2022}), and its
greatest weakness. The abstract nature of the Price equation places it
at the top of the hierarchy of fundamental theorems of evolution from
which the rest (Robertson's theorem, Fisher's fundamental theorem,
breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be
easily derived (\citeproc{ref-Lehtonen2016}{Lehtonen 2016},
\citeproc{ref-Lehtonen2018}{2018}; \citeproc{ref-Queller2017}{Queller
2017}). This abstractness is also key to developing a more general view
of evolution (\citeproc{ref-Rice2020}{Rice 2020};
\citeproc{ref-Luque2021}{Luque and Baravalle 2021};
\citeproc{ref-Edelaar2023}{Edelaar et al. 2023}). In contrast, some
researchers consider the Price equation just a triviality (even
tautological), and useless without further modelling assumptions
(\citeproc{ref-VanVeelen2005}{van Veelen 2005};
\citeproc{ref-VanVeelen2012}{van Veelen et al. 2012}). The debate
remains open (\citeproc{ref-VanVeelen2020}{van Veelen 2020};
\citeproc{ref-Baravalle2024}{Baravalle et al. 2024}).
\end{quote}

@@

\begin{quote}
\textbf{Box 2:} The number of individuals in any closed population
(\(N\)) at any given time (\(t + 1\)) is determined by the existing
number of individuals (\(N_{t}\)), plus the number of births
(\(Births\)) minus the number of deaths (\(Deaths\)),
\[N_{t+1} = N_{t} + Births - Deaths.\] This equation is necessarily true
for any closed population. Despite its simplicity, it is a general
equation for defining population change and a starting point for
understanding population ecology. Turchin
(\citeproc{ref-Turchin2001}{2001}) notes that a consequence of this
fundamental equation is the tendency for populations to grow
exponentially (technically geometrically in the above case where time is
discrete). This inherent underlying tendency towards exponential growth
persists even as the complexities of real populations, such as
structure, stochasticity, or density-dependent effects are added to
population models (\citeproc{ref-Turchin2001}{Turchin 2001}). Given the
assumption that all individuals in the population are identical, a per
capita rate of birth, \(Births = bN_{t}\) and death, \(Deaths = dN_{t}\)
can be defined. Rearranging and defining \(\lambda = 1 + b - d\) gives,
\(N_{t+1} = N_{t}\lambda\). Here \(\lambda\) is the finite rate of
increase (\citeproc{ref-Gotelli2001}{Gotelli 2001}), and note that
because \(0 \leq d \leq 1\), \(\lambda \geq 0\). Verbally, the change in
size of any closed population equals its existing size times its finite
rate of increase.
\end{quote}

@@

To fully unify biological evolution and population ecology, we must
reconcile the the Price equation (Box 1) with the general equation for
population change (Box 2). The Price equation is critical for
partitioning different components of biological change
(\citeproc{ref-Price1970}{Price 1970}; \citeproc{ref-Frank1997}{Frank
1997}; \citeproc{ref-Gardner2008}{Gardner 2008};
\citeproc{ref-Luque2016}{Luque 2017}; \citeproc{ref-Queller2017}{Queller
2017}; \citeproc{ref-Lehtonen2018}{Lehtonen 2018}). It has also been
highly useful for integrating evolutionary theory across disciplines
(\citeproc{ref-Fox2006}{Fox 2006};
\citeproc{ref-Brantingham2007}{Brantingham 2007};
\citeproc{ref-MacCallum2012}{MacCallum et al. 2012};
\citeproc{ref-Frank2015}{Frank 2015}; \citeproc{ref-Godsoe2021}{Godsoe
et al. 2021}; \citeproc{ref-Ulrich2024}{Ulrich et al. 2024}). These
properties would seem to make it an intuitive starting point for a
logical foundation of ecology and evolution, perhaps through some kind
of mathematical equivalence (\citeproc{ref-Page2002}{Page and Nowak
2002}) or addition of terms (\citeproc{ref-Collins2009}{Collins and
Gardner 2009}), or through the use of its recursive structure
(\citeproc{ref-Kerr2009}{Kerr and Godfrey-Smith 2009};
\citeproc{ref-Frank2012}{Frank 2012}). But despite its flexibility, the
Price equation still relies on relative frequencies, which must by
definition sum to one (\citeproc{ref-Frank2015}{Frank 2015}). This is
because the Price equation describes the average change in a population;
the frequency of entities is scaled thereby conserving total probability
(\citeproc{ref-Frank2015}{Frank 2015}, \citeproc{ref-Frank2016}{2016}).
But to recover the fundamental principle of exponential population
growth (\citeproc{ref-Turchin2001}{Turchin 2001}), this scaling must be
avoided in a fundamental equation of ecology and evolution.

@@

From eqn 1, we can derive the most fundamental equations of population
ecology (Box 2) and evolutionary biology (Box 1) through an appropriate
interpretation of \(z_{i}\).

@@

To recover the general equation for population ecology (Box 2), we
define \(z_{i}\) as the identity of \(i\) belonging to the population.
In other words, we set \(z_{i} = 1\) to simply indicate that \(i\) is a
member of the population. Further, we assume that individuals do not
change species by setting \(\Delta z_{i} = 0\). In this case,

@@

If we define \(\lambda = 1 + \beta - \delta\) (Box 2), then we can
rewrite,

@@

From eqn 1, which describes fundamental birth and death processes in a
population, we can derive both the most fundamental model of population
ecology (eqn 2; Box 2) and the fundamental equation of evolution (eqn
10; Box 1). We might also consider \(\Omega\) to be a quantity of
ecosystem function, in that it is the summed effect of the traits
(\(z_{i}\)) of individual entities (\(i\)) in the population. For
example, \(z\) might be interpreted as biomass or photosynthetic rate.

@@

Rather than making simplifying assumptions, as is the approach for most
ecological and evolutionary models, we focus on fundamental axioms that
are universal to closed biological systems: discrete individuals, birth,
death, and change over time. We define an abstract sum (\(\Omega\)), to
which all individuals within the population contribute. From the simple
assumptions of population identity (\(z_{i} = 1\)) and invariability
(\(\Delta z_{i} = 0\)), we recover the most general equation of
population ecology (Box 2) and principle of exponential growth
(\(N_{t+1} = N_{t}\lambda\)). By defining individual fitness (\(w_{i}\))
and applying the total conservation of probability to individual
frequencies (\citeproc{ref-Frank2015}{Frank 2015},
\citeproc{ref-Frank2016}{2016}), we recover the most fundamental
equation of evolution (Box 1). We therefore deliver a general framework
to unify fundamental equations of ecological and evolutionary change.

@@

Our unification demonstrates the equivalence between the finite rate of
increase \(\lambda\) (Box 2) (\citeproc{ref-Gotelli2001}{Gotelli 2001})
and population mean evolutionary fitness \(\bar{w}\) (Box 1). The
population growth equation \(N_{t+1} = N_{t}\lambda\) could therefore be
rewritten as \(N_{t+1} = N_{t}\bar{w}\). We do not claim to be the first
to notice this equivalence between population growth and mean population
fitness. Indeed, the relationship between population growth rate and
evolutionary fitness has been proposed and applied many times before
(\citeproc{ref-Fisher1930}{Fisher 1930};
\citeproc{ref-Charlesworth1980}{Charlesworth 1980};
\citeproc{ref-Lande1982}{Lande 1982}; \citeproc{ref-Roff2008}{Roff
2008}; \citeproc{ref-Lion2018}{Lion 2018}). For example, Lande
(\citeproc{ref-Lande1982}{1982}) explicitly concludes mean absolute
fitness per unit time is \(\bar{w} = e^{r}\), where \(r = ln(\lambda)\).
We show this from first principles and clarify the relationship between
fitness and population growth. Over an arbitrary length of time, fitness
is properly defined as \(w_{i} = \beta_{i} - \delta_{i} + 1\). Over an
individual's lifetime (which, by definition, includes death), fitness is
therefore \(\beta_{i}\). Interestingly, the rate of change in ecology
and evolution are reflected in the first and second statistical moments
of fitness, respectively. Population growth rate reflects mean fitness
\(\bar{w}\), while the rate of evolutionary change reflects the variance
in fitness \(\mathrm{Var}(w)\) (i.e., Fisher's fundamental theorem)
(\citeproc{ref-Frank1997}{Frank 1997}; \citeproc{ref-Rice2004}{Rice
2004}; \citeproc{ref-Queller2017}{Queller 2017}).

@@

We have focused on the dynamics of a closed population, and in doing so
leave ecological and evolutionary change attributable to migration for
future work. In population ecology, immigration and emigration can be
incorporated by adding a term for each to the right-hand side of the
equation in Box 2 (\citeproc{ref-Gotelli2001}{Gotelli 2001}). In
evolution, because the Price equation relies on mapping
ancestor-descendant relationships, accounting for migration is more
challenging. Kerr and Godfrey-Smith (\citeproc{ref-Kerr2009}{2009})
demonstrate how the Price equation can be extended to allow for
arbitrary links between ancestors and descendants, thereby extending the
Price equation to allow for immigration and emigration. Frank
(\citeproc{ref-Frank2012}{2012}) presents a simplified version of Kerr
and Godfrey-Smith (\citeproc{ref-Kerr2009}{2009}) that allows some
fraction of descendants to be unconnected to ancestors. In both ecology
and evolution, accounting for migration is done through the use of
additional terms on the right-hand side of the equations.

Our fundamental equation is complete and exact. It therefore implicitly
includes any effects of density dependence on population growth (see Box
2), or any social effects on evolutionary change (see Box 1). We can
make both of these effects explicit with the same partition in the
summation on the right-hand side of eqn 1. To do this, we multiply the
expression
\(\left(\beta_{i} - \delta_{i} + 1 \right)\left(z_{i} + \Delta z_{i} \right)\)
by
\(\left(1 - \sum_{j=1}^{N} a_{ij}\left(z_{j} + \Delta z_{j}\right)\right)\),
where \(j\) indexes the same individuals as \(i\), and \(a_{ij}\)
defines the effect \(j\) has on the contribution of \(i\) to \(\Omega\).
In evolutionary terms, \(a_{ij}\) defines how \(j\) modulates the
fitness of \(i\) (\(w_{i}\)). In ecological terms, \(a_{ij}\) defines
how \(j\) modulates the contribution of \(i\) to total population
growth. If we define \(\alpha\) as the summed effect of all individuals
on a focal \(i\) (i.e., \(\alpha = \sum_{j=1}^{N} a_{ij}\)), then we can
partition ecological change into density-independent and
density-dependent effects and derive a discrete time logistic growth
equation. We save these derivations for a future investigations focused
on evolutionary and community ecology.
